---
title: "Fallstudie Basecode Tutorial - Replication"
author: "Daniel Zimmermann / David Lilek"
output:
  html_document:
    df_print: paged
    fig.align: center
    self_contained: yes
    fig.height: 4
    fig.width: 8
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: no
    code_folding: hide
---

# Remark

This file contains a reproduction and discussion of the guided clustering tutorial. Theory behind scRNA-seq and the methods used here can be found in the written report.

# Aim

The aim was to reproduce the [guided clustering tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html) provided by Seurat. 

# Setting up the environment

As stated in the written report the [installation of Seurat](https://satijalab.org/seurat/articles/install.html) was straightforward. With the command `sessionInfo()`it is easy to check whether the correct versions for all packages are installed on the used (local) machine.

# Load libraries

```{r message=FALSE, warning=FALSE}
###################
#load libraries
###################
library(dplyr)
library(Seurat)
library(patchwork)
```

# Setup the Seurat Object

For loading and setting up the Seurat object the `data.dir`has to be set correctly. This folder contains three files:

* barcodes.tsv: contains the barcodes for the reads
* genes.tsv: contains 2 columns
  + column1: ensemble identifiers for the genes resp ensemble gene record (ENSG)    + column2: [gene name](https://en.wikipedia.org/wiki/Gene_nomenclature)
  + [Link](https://www.ebi.ac.uk/training/online/courses/ensembl-browsing-genomes/navigating-ensembl/investigating-a-gene/)
* matrix.mtx: 
  + contains the cellular barcodes 
  + "Each element of the matrix is the number of UMIs associated with a feature (row) and a barcode (column)"
  + [Link](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/latest/output/matrices)

The function `Read10X()` then reads in the output from the cellranger 10x pipeline. It returns a unique molecular identified (UMI) count matrix where the values represent the "molecules for each feature (i.e. gene; row) that are detected in each cell (column)". This data is then converted into a `Seurat` object.


```{r message=FALSE, warning=FALSE}
############################
#Setup the Seurat Object
############################

# Load the PBMC dataset
pbmc.data <- Read10X(data.dir = "./data/")
# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc
```

# Standard pre-processing workflow

## QC and selecting cells for further analysis

* Visualize QC metrics as a violin plot
* Based on these plots, the following criteria are chosen:
  + 200 < nFeature_RNA < 2000
  + percent.mt < 5 %

```{r}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
# Visualize QC metrics as a violin plot
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

The `FeatureScatter()` function is a x,y scatter plot of feature-feature relationships. It can be seen that there is a strong correlation between nFeature_RNA and nCount_RNA, while no correlation between percent.mt and nCount_RNA/nFeature_RNA exists.

```{r}
# FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.

plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

# Filtering

As mentioned before, only cells with a unique feature count between 200 and 2,500 as well as a percentage of mitochondrial RNA less than 5 % are used for further analysis. This is achieved using the `subset()` function.

```{r}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```


# Normalizing the data

For normalization the "LogNormalize" method is used. First, expression levels are normalized per cell by dividing the number of reads for each feature by the total number of reads. The results are scaled by a factor of 10,000 and finally log transformed (ln(x+1)).

```{r}
# pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
pbmc <- NormalizeData(pbmc)
```

# Identification of highly variable features (feature selection)

By default 2,000 features are returned by the function `FindVariableFeatures()` which are then used for downstream analysis. The selection is based on the count data and calculating the variance-to-mean ratio. The red dots in the figure below represent features which are selected for further analysis. The 10 most highly variable genes are labelled additionally in the right plot.

```{r fig.width=12}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

# Scaling the data

For many data analysis steps, such as PCA, the data has to be scaled first. In Seurat, scaling is by default done by mean-centering the data and scaling to unit variance.

```{r}
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
```

# Perform linear dimensional reduction

For dimension reduction a PCA is performed on the scaled data. The PC'S are "representing a ‘metafeature’ that combines information across a correlated feature set" as can be seen in the loading-plots. The loadings plots show how the features are weighted per PC.
`DimHeatmap()`can be used for further exploration of the (heterogeneity in the) data-set and is suitable to select the most important PCs. The rows of the heatmap shows the features. The columns represent the single cells. Both are sorted by their PC scores. In this case especially in the first two PCs quite high heterogeneity can be seen. Some features are either present in one group or in the other group of cells.
This can be also seen in the `DimPlot` - in this case PC1 vs PC2 is plotted - where some separation between the different cells can be seen.

```{r}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
# Examine and visualize PCA results a few different ways
print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
DimPlot(pbmc, reduction = "pca")
DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)
DimHeatmap(pbmc, dims = 1:15, cells = 500, balanced = TRUE)
```

# Determine the ‘dimensionality’ of the dataset

For the clustering approach afterwards only the most important PCs should be used. To determine those PCs Seurat offers 2 possibilities. 

* JackStraw procedure
  + random permutation of a subset
  + for this sub-set PCA scores are calculated and compared to the "original" PCA score 
  + p value is calculated based on gene association with each PC
* Ellbow plot - standard deviation of each PC is plotted 

Especially the JackStraw procedure is quite time consuming and the user can select different options - e.g. how man replicates are performed. 

As can be seen in the JackStraw plot 13 PCs should be used. The elbow plot suggests that app. 10 PCs are appropriate. Combining those findings with die heatmap from before 10 PCs were used.

But as stated in the tutorial also other things have to be taken into account and also a comparison of downstream analysis results using different number of PCs should be performed. They also recommend to "err on the higher side" if the number of PCs is chosen.

```{r}
# NOTE: This process can take a long time for big datasets, comment out for expediency. More
# approximate techniques such as those implemented in ElbowPlot() can be used to reduce
# computation time
pbmc <- JackStraw(pbmc, num.replicate = 100)
pbmc <- ScoreJackStraw(pbmc, dims = 1:20)
JackStrawPlot(pbmc, dims = 1:15)
ElbowPlot(pbmc)
```

# Cluster the cells

For clustering Seurat uses a graph-based approach using the functions `FindNeighbors` and `FindClusters`.
In the `FindNeighbors` the number of used PCs can be defined using the dims parameter - in this case PC1-PC10 were chosen. The function constructs a "KNN graph based on the euclidean distance in PCA space". The edge weights are then refined.
In the `FindClusters` function - modularity optimization technique using Louvain algorithm (default) or SLM  algorithm - the resolution is the most important parameter. By increasing this parameter also the number of clusters will increase. Seurat recommends values between 0.4-1.2 for app. 3000 cells. The resolution also increases if the number of cells increases.
The `ident` function can be used to show the cluster IDs.

```{r}
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)
# Look at cluster IDs of the first 5 cells
head(Idents(pbmc), 5)
```

# Run non-linear dimensional reduction (UMAP/tSNE)

Seurat uses non-linear dimensional reduction techniques (tSNE and UMAP are available) for further visualization and exploration of the data. Seurat recommends to use the same number of PCs for clustering and dimensional reduction. As can be seen 8 different clusters are formed independent if UMAP or tSNE are used.

Compared to the result in the tutorial the results from UMAP differs a bit. Here are 2 statements which we found during our online search:
* [LINK1](https://stackoverflow.com/questions/67101829/seurat-umap-visualization-result-is-mirrored-after-running-in-two-identical-envi): "The exact location of points on a UMAP plot can chance across different computers and OSs. We do our best to minimize any randomness to the procedure by fixing the random seed, but some fluctuation across systems is inevitable, and nothing to worry about.
* [LINK2](https://github.com/satijalab/seurat/issues/536)"I wouldn't expect UMAP to be able to pick up R's seed, as these are different environments. Running UMAP within R uses the reticulate package to call UMAP's Python code from within R. Reticulate doesn't automatically pass R's random seed to Python, so there's no reason to expect UMAP to pick up R's random seed. However, this would be a good thing to allow users to set. As for the issue with random_state, that seems to be an issue with UMAP itself, not Seurat."

This can result in slightly different UMAP results. What we found out that using the latest Seurat version it gives the same results, also for other students.


* UMAP
```{r}
# If you haven't installed UMAP, you can do so via reticulate::py_install(packages =
# 'umap-learn')
pbmc <- RunUMAP(pbmc, dims = 1:10)
# note that you can set `label = TRUE` or use the LabelClusters function to help label
# individual clusters
DimPlot(pbmc, reduction = "umap")
#saveRDS(pbmc, file = "./pbmc_tutorial.rds")
```

* tSNE

```{r}
#run TSNE
pbmc <- RunTSNE(pbmc, dims = 1:10)
TSNEPlot(pbmc)
```

# Finding differentially expressed features (cluster biomarkers)

To find markers (`FindMarkers`) which define a certain cluster DE (differential expression) is used. The parameter ident.1 defines the comparison/identification of cluster markers (positive or negative) compared to the other cells - e.g. ident.1 = 2 means that cluster2 is used - see below.  [LINK](https://satijalab.org/seurat/reference/findmarkers) 
A more complex example is to find markers distinguishing cluster 5 from cluster 0/2. This can be done using the command `FindMarkers(pbmc, ident.1 = 5, ident.2 = c(0, 3), min.pct = 0.25)` as shown below.
Also an example to find markers for every cluster compared to all remaining cells and report only the positive ones is shown below.

Setting the `min.pct` argument -  "min.pct argument requires a feature to be detected at a minimum percentage in either of the two groups of cells" - to 0 will increase the running time dramatically.

* find all markers of cluster 2
```{r}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(pbmc, ident.1 = 2, min.pct = 0.25)
head(cluster2.markers, n = 5)
```

* find all markers distinguishing cluster 5 from clusters 0 and 3

```{r}
# find all markers distinguishing cluster 5 from clusters 0 and 3
cluster5.markers <- FindMarkers(pbmc, ident.1 = 5, ident.2 = c(0, 3), min.pct = 0.25)
head(cluster5.markers, n = 5)
```

* find markers for every cluster compared to all remaining cells, report only the positive ones

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive
# ones
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
pbmc.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC)
```

* For testing DE Seurat has several options which can be defined via the parameter `test.use`. In this case ROC test, which "returns the 'classification power' for any individual marker" was chosen.

```{r}
cluster0.markers <- FindMarkers(pbmc, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)
VlnPlot(pbmc, features = c("MS4A1", "CD79A"))

```

* Also the plotting of the raw counts is possible.

```{r}
# you can plot raw counts as well
VlnPlot(pbmc, features = c("NKG7", "PF4"), slot = "counts", log = TRUE)
```

* For visualization the marker expression violine plots are used using the `FeaturePlot()` function.

```{r}
FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP",
    "CD8A"))
```

* DoHeatmap
  + The `DoHeatmap()` function generates a heatmap for cells (rows) and features (columns). In this case the top10 markers for each clusters are used.
  + As expected it can be seen that in each cluster different markers are present.
  + Remark: in the online tutorial from Seurat they wrote: "DoHeatmap() generates an expression heatmap for given cells and features. In this case, we are plotting the top 20 markers (or all markers if less than 20) for each cluster." -> BUT: in the code from the tutorial they only used 10 markers as in the code below


```{r}
pbmc.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(pbmc, features = top10$gene) + NoLegend()
```

# Assigning cell type identity to clusters

In the last step markers can be manually assigned to cell types and used for labelling the clusters in UMAP.

```{r}
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono",
    "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```

